---
title: "Tamas_Toth_MSDS_6372_Project1"
author: "Tamas Toth"
date: '2022-05-27'
output:
  #html_document:
    #theme: cerulean
    #highlight: textmate
  github_document:
  toc: FALSE
  toc_depth: 3
  fig_width: 7
  fig_height: 5
  dev: "png"
  df_print: "default"
  includes: NULL
  md_extensions: NULL
  hard_line_breaks: TRUE
  pandoc_args: NULL
  html_preview: TRUE
  keep_html: TRUE
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#### Loading the necessary R libraries for the analysis
```{r message = FALSE}
# Load the necessary libraries
library(knitr)
library(rmarkdown)
library(ggpubr)
library(dplyr)
library(tidyr)
library(plyr)
library(ggplot2)
library(tidyverse)
library(ggthemes)
library(e1071)
library(class)
library(caret)
library(stringr)
library(sjPlot)
library(data.table)
library(reshape2)
library(corrplot)
library(naivebayes)
library(car)
library(egg)
library(rworldmap)
library(Hmisc)
library(DataExplorer)
library(selectiveInference)
library(dlookr)
```
```{r message = FALSE}
# Turn off scientific notation
options(scipen = 100, digits = 4)
```
#### Read the data
```{r,fig.align='center',out.extra='angle=90', message = FALSE}
#Read the data
#setwd('/Users/ttoth76/Downloads/SMU/Semester_2/DS 6372 Applied Statistics_Inference & Modeling/FLS/Project1_Summer2022/GitContent/LifeExpectancy')
LifeExp = read.csv(file = 'Life_Expectancy_Data.csv',header = TRUE, sep = ",")
# take a sample of 15 from the dataframe
LifeExp_sample = sample_n(LifeExp, 5)
knitr::kable(LifeExp_sample, "html")
```

#### Address the missing values in each column (NA as well as empty strings).
```{r}
# Address the missing values in each column (NA as well as empty strings).
missing_df = as.data.frame(sapply(LifeExp, function(x) sum(is.na(x))))
colnames(missing_df) = c("variable missing")
knitr::kable(missing_df, "html")
empty_string_df = as.data.frame(sapply(LifeExp, function(x) sum(x == "")))
colnames(empty_string_df) = c("variable empty")
knitr::kable(empty_string_df, "html")
```

```{r}
#set random seed
set.seed(329)
```

```{r, warning=FALSE}
# Function to Identify different characteristics of the data frame 
# Getting a concise summary of the dataframe: str()
# Listing the column labels of the dataframe: colnames()
# Size of the dataset: dim()
# # Verify if there is any negative values in the dataset
dfinfo = function(df_name)
  {
  df_structure = str(df_name)
  df_colnames = colnames(df_name)
  df_dimensions = dim(df_name)
  df_neg = print(paste("Negative values in the Data Frame:", 
                       sapply(df_name, function(x) sum(x < 0))))
  outparam = list(df_structure, df_colnames, df_dimensions, df_neg)
  return (outparam)
}
```
```{r}
dfinfo(LifeExp)
```


#### Generate summary statistics
```{r}
# Generate summary statistics
summary(LifeExp)
```

### Observations:
* The dataset is comprised of 2938 observations and 22 variables
* There are numerical and categorical variables (Country and Status) in the dataset
* Column names have spaces and special characters that has been replaced by R with "."
* There are missing values or empty strings in the dataset
* ????? duplicated records????????
* 'Life.expectancy' is the dependent variable - There are 10 missing observations in the dependent variable
* We need to predict Salary however there is no salary variable in the dataset but MonthlyIncome variable seems to be sufficient for this purpose.


### Scatterplots
```{r Linear regression}
#####################################################################################
#                        Scatter plots for checking linearity                       #
#####################################################################################

################### Linear - Linear ###################
LifeExp$Status = as.factor(LifeExp$Status)
#num_cols = LifeExp %>% select(where(is.numeric)) %>% colnames()
#MLR_num_LE = LifeExp[, num_cols]
pairs(Life.expectancy~Year+Adult.Mortality+infant.deaths+Alcohol, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))
pairs(Life.expectancy~log(percentage.expenditure)+Hepatitis.B+Measles+BMI, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))
pairs(Life.expectancy~under.five.deaths+Polio+Total.expenditure+Diphtheria, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))
pairs(Life.expectancy~HIV.AIDS+GDP+Population+thinness..1.19.years, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))
pairs(Life.expectancy~thinness.5.9.years+Income.composition.of.resources+Schooling, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))


################### Linear - Log transformation ###################
LifeExp$Status = as.factor(LifeExp$Status)
#num_cols = LifeExp %>% select(where(is.numeric)) %>% colnames()
#MLR_num_LE = LifeExp[, num_cols]
pairs(Life.expectancy~Year+Adult.Mortality+infant.deaths+Alcohol, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))
pairs(Life.expectancy~log(percentage.expenditure)+Hepatitis.B+Measles+BMI, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))
pairs(Life.expectancy~under.five.deaths+Polio+Total.expenditure+Diphtheria, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))
pairs(Life.expectancy~log(HIV.AIDS)+log(GDP)+Population+thinness..1.19.years, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))
pairs(Life.expectancy~thinness.5.9.years+Income.composition.of.resources+Schooling, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))

################### Log - Log transformation ###################
LifeExp$Status = as.factor(LifeExp$Status)
#num_cols = LifeExp %>% select(where(is.numeric)) %>% colnames()
#MLR_num_LE = LifeExp[, num_cols]
pairs(log(Life.expectancy)~Year+Adult.Mortality+infant.deaths+Alcohol, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))
pairs(log(Life.expectancy)~log(percentage.expenditure)+Hepatitis.B+Measles+BMI, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))
pairs(log(Life.expectancy)~under.five.deaths+Polio+Total.expenditure+Diphtheria, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))
pairs(log(Life.expectancy)~log(HIV.AIDS)+log(GDP)+Population+thinness..1.19.years, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))
pairs(log(Life.expectancy)~thinness.5.9.years+Income.composition.of.resources+Schooling, data=LifeExp, col=ifelse(LifeExp$Status=="Developed", "green", "red"))



```
### Observations
We can observe linear relationship between Life Expectancy and the following variables

* Income composition of resources
* Schooling
* Log(HIV.AIDS)
* Log(GDP)
* Log(percentage.expenditure)
* BMI
* Year
* Adult.Mortality

#adding region column to do regional imputation
```{r adding region column}
###rename ivory coast and remove parenthesis from country names
LifeExp$Country <- replace(LifeExp$Country,LifeExp$Country == "Côte d'Ivoire","Ivory Coast")
LifeExp$Country <-  str_replace(LifeExp$Country,"\\(.*\\)","")
LifeExp$Country <- trimws(LifeExp$Country,"right")

### add region using rworldmap package
regions <- rworldmap::countryRegions
regions <- regions[,c(2,6)]
colnames(regions) <- c("Country","Region")
regions$Country <- replace(regions$Country, regions$Country == "The Bahamas", "Bahamas")
regions$Country <- replace(regions$Country, regions$Country == "Brunei", "Brunei Darussalam")
regions$Country <- replace(regions$Country, regions$Country == "Cape Verde", "Cabo Verde")
regions$Country <- replace(regions$Country, regions$Country == "Republic of the Congo", "Congo")
regions$Country <- replace(regions$Country, regions$Country == "Czech Republic", "Czechia")
regions$Country <- replace(regions$Country, regions$Country == "North Korea", "Democratic People's Republic of Korea")
regions$Country <- replace(regions$Country, regions$Country == "Guinea Bissau", "Guinea-Bissau")
regions$Country <- replace(regions$Country, regions$Country == "Laos","Lao People's Democratic Republic")
regions$Country <- replace(regions$Country, regions$Country == "Federated States of Micronesia", "Micronesia")
regions$Country <- replace(regions$Country, regions$Country == "South Korea", "Republic of Korea")
regions$Country <- replace(regions$Country, regions$Country == "Moldova", "Republic of Moldova")
regions$Country <- replace(regions$Country, regions$Country == "Russia", "Russian Federation")
regions$Country <- replace(regions$Country, regions$Country == "Republic of Serbia", "Serbia")
regions$Country <- replace(regions$Country, regions$Country == "Syria", "Syrian Arab Republic")
regions$Country <- replace(regions$Country, regions$Country == "Macedonia", "The former Yugoslav republic of Macedonia")
regions$Country <- replace(regions$Country, regions$Country == "East Timor", "Timor-Leste")
regions$Country <- replace(regions$Country, regions$Country == "United Kingdom", "United Kingdom of Great Britain and Northern Ireland")
regions$Country <- replace(regions$Country, regions$Country == "Vietnam", "Viet Nam")


LifeExp <-  join(LifeExp,regions,by = "Country", type = 'left')
LifeExp$Country <- replace(LifeExp$Country, LifeExp$Country == "Micronesia", "Micronesia (Federated States of)")
LifeExp$Region <- as.factor(LifeExp$Region)
```


## Fixing the missing values by replacing with median
```{r median imputation}
# Drop missing values from the dependent variable
LifeExp = LifeExp[!(is.na(LifeExp$Life.expectancy)),]

na_list = colnames(LifeExp)[apply(LifeExp, 2, anyNA)]
# LifeExp_db = apply(LifeExp[,colnames(LifeExp) %in% na_list],2,median,na.rm =  TRUE)
LifeExp = LifeExp %>% group_by(Country) %>% dplyr::mutate(Alcohol = replace(Alcohol,is.na(Alcohol), median(Alcohol, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Country) %>% dplyr::mutate(Hepatitis.B = replace(Hepatitis.B,is.na(Hepatitis.B), median(Hepatitis.B, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Country) %>% dplyr::mutate(BMI = replace(BMI,is.na(BMI), median(BMI, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Country) %>% dplyr::mutate(Polio = replace(Polio,is.na(Polio), median(Polio, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Country) %>% dplyr::mutate(Total.expenditure = replace(Total.expenditure,is.na(Total.expenditure), median(Total.expenditure, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Country) %>% dplyr::mutate(Diphtheria = replace(Diphtheria,is.na(Diphtheria), median(Diphtheria, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Country) %>% dplyr::mutate(GDP = replace(GDP,is.na(GDP), median(GDP, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Country) %>% dplyr::mutate(Population = replace(Population,is.na(Population), median(Population, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Country) %>% dplyr::mutate(thinness..1.19.years = replace(thinness..1.19.years,is.na(thinness..1.19.years), median(thinness..1.19.years, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Country) %>% dplyr::mutate(thinness.5.9.years = replace(thinness.5.9.years,is.na(thinness.5.9.years), median(thinness.5.9.years, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Country) %>% dplyr::mutate(Income.composition.of.resources = replace(Income.composition.of.resources,is.na(Income.composition.of.resources), median(Income.composition.of.resources, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Country) %>% dplyr::mutate(Schooling = replace(Schooling,is.na(Schooling), median(Schooling, na.rm = TRUE)))

#repeat grouping by Region for countries which have only NAs
LifeExp = LifeExp %>% group_by(Region) %>% dplyr::mutate(Alcohol = replace(Alcohol,is.na(Alcohol), median(Alcohol, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Region) %>% dplyr::mutate(Hepatitis.B = replace(Hepatitis.B,is.na(Hepatitis.B), median(Hepatitis.B, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Region) %>% dplyr::mutate(BMI = replace(BMI,is.na(BMI), median(BMI, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Region) %>% dplyr::mutate(Polio = replace(Polio,is.na(Polio), median(Polio, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Region) %>% dplyr::mutate(Total.expenditure = replace(Total.expenditure,is.na(Total.expenditure), median(Total.expenditure, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Region) %>% dplyr::mutate(Diphtheria = replace(Diphtheria,is.na(Diphtheria), median(Diphtheria, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Region) %>% dplyr::mutate(GDP = replace(GDP,is.na(GDP), median(GDP, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Region) %>% dplyr::mutate(Population = replace(Population,is.na(Population), median(Population, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Region) %>% dplyr::mutate(thinness..1.19.years = replace(thinness..1.19.years,is.na(thinness..1.19.years), median(thinness..1.19.years, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Region) %>% dplyr::mutate(thinness.5.9.years = replace(thinness.5.9.years,is.na(thinness.5.9.years), median(thinness.5.9.years, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Region) %>% dplyr::mutate(Income.composition.of.resources = replace(Income.composition.of.resources,is.na(Income.composition.of.resources), median(Income.composition.of.resources, na.rm = TRUE)))
LifeExp = LifeExp %>% group_by(Region) %>% dplyr::mutate(Schooling = replace(Schooling,is.na(Schooling), median(Schooling, na.rm = TRUE)))
# convert the tibble to data frame
LifeExp = as.data.frame(LifeExp)

#impute USA values
#USA GDP
US_GDP <- read.csv('API_NY.GDP.PCAP.CD_DS2_en_csv_v2_4150786.csv',header = FALSE)
colnames(US_GDP) <- US_GDP[3,]
US_GDP <- rename(US_GDP,c("Country Name"="Country"))
US_GDP <- US_GDP[US_GDP$Country == "United States",c(1,45:60)]
US_GDP <- US_GDP %>% pivot_longer(!Country,names_to = "Year",values_to = "GDP2")
US_GDP$Country <- replace(US_GDP$Country, US_GDP$Country == "United States", "United States of America")
US_GDP$Year <- as.integer(US_GDP$Year)

LifeExp <- left_join(LifeExp,US_GDP, by=c("Country","Year"))
LifeExp <- LifeExp %>% dplyr::mutate(GDP = ifelse(LifeExp$Country == "United States of America", LifeExp$GDP2, LifeExp$GDP))

#US Schooling
US_Scho <- read.csv("Expected years of schooling (years).csv",skip = 6,header = FALSE)
colnames(US_Scho) <- US_Scho[1,]
US_Scho$Country <- trimws(US_Scho$Country, which = "both")
US_Scho <- US_Scho[US_Scho$Country == "United States", colSums(is.na(US_Scho)) !=nrow(US_Scho)]
US_Scho <- US_Scho[,c(2,13:28)]
US_Scho <- US_Scho %>% pivot_longer(!Country,names_to = "Year",values_to = "Schooling2")
US_Scho$Year <- as.integer(US_Scho$Year)
US_Scho$Schooling2 <- as.numeric(US_Scho$Schooling2)
US_Scho$Country <- replace(US_Scho$Country, US_Scho$Country == "United States", "United States of America")


LifeExp <- left_join(LifeExp,US_Scho,by=c("Country","Year"))
LifeExp <- LifeExp %>% dplyr::mutate(Schooling =ifelse(LifeExp$Country == "United States of America",LifeExp$Schooling2,LifeExp$Schooling))

#US income composition
US_Inc <- read.csv('Income index.csv',skip = 5, header = FALSE)
colnames(US_Inc) <- US_Inc[1,]
US_Inc$Country <- trimws(US_Inc$Country,which = "both")
US_Inc <- US_Inc[US_Inc$Country == "United States",colSums(is.na(US_Inc)) !=nrow(US_Inc)]
US_Inc <- US_Inc[,c(2,13:28)]
US_Inc <- US_Inc %>%  pivot_longer(!Country,names_to = "Year", values_to = "comp2")
US_Inc$Year <- as.integer(US_Inc$Year)
US_Inc$comp2 <- as.numeric(US_Inc$comp2)
US_Inc$Country <- replace(US_Inc$Country,US_Inc$Country == "United States", "United States of America")

LifeExp <- left_join(LifeExp,US_Inc,by=c("Country","Year"))
LifeExp <- LifeExp %>% dplyr::mutate("Income.composition.of.resources" = ifelse(LifeExp$Country == "United States of America", LifeExp$comp2,LifeExp$Income.composition.of.resources))

#Add population from external source
#data from UN in thousands 
pop_all <- read.csv('WPP2019_TotalPopulationBySex.csv')
pop_all <- pop_all %>% dplyr::select(Country = Location, Year = Time, Population2 = PopTotal) %>% mutate(Population2 = Population2*1000) %>% filter(Year %in% c(2000:2015))

#clean country names to match before merge
pop_all$Country <- replace(pop_all$Country, pop_all$Country == "Bolivia (Plurinational State of)", "Bolivia")
pop_all$Country <- replace(pop_all$Country,pop_all$Country == "Côte d'Ivoire", "Ivory Coast")
pop_all$Country <- replace(pop_all$Country,pop_all$Country == "Dem. People's Republic of Korea","Democratic People's Republic of Korea")
pop_all$Country <- replace(pop_all$Country, pop_all$Country == "Iran (Islamic Republic of)", "Iran")
pop_all$Country <- replace(pop_all$Country, pop_all$Country == "Micronesia (Fed. States of)", "Micronesia (Federated States of)")
pop_all$Country <- replace(pop_all$Country, pop_all$Country == "Eswatini", "Swaziland")
pop_all$Country <- replace(pop_all$Country, pop_all$Country == "North Macedonia", "The former Yugoslav republic of Macedonia")
pop_all$Country <- replace(pop_all$Country, pop_all$Country == "United Kingdom", "United Kingdom of Great Britain and Northern Ireland")
pop_all$Country <- replace(pop_all$Country, pop_all$Country == "Venezuela (Bolivarian Republic of)", "Venezuela")
LifeExp<- left_join(LifeExp,pop_all,by=c("Country","Year")) %>% mutate(Population = Population2)

#replace adult.mortality since there are clear mistakes with the data that could
#not be resolved with transformation of the variable
adlt_mort <- read.csv('Adult_mort.csv',header = TRUE)
adlt_mort <- adlt_mort %>% dplyr::select("Country" = Location,"Year" = Period,"Adult.Mort2" = Value)
adlt_mort$Country <- replace(adlt_mort$Country,adlt_mort$Country == "Côte d’Ivoire","Ivory Coast")
adlt_mort$Country <- replace(adlt_mort$Country,adlt_mort$Country == "Bolivia (Plurinational State of)","Bolivia")
adlt_mort$Country <- replace(adlt_mort$Country,adlt_mort$Country == "Venezuela (Bolivarian Republic of)","Venezuela")
adlt_mort$Country <- replace(adlt_mort$Country,adlt_mort$Country == "Iran (Islamic Republic of)", "Iran")
adlt_mort$Country <- replace(adlt_mort$Country,adlt_mort$Country == "Eswatini", "Swaziland")
adlt_mort$Country <- replace(adlt_mort$Country,adlt_mort$Country == "The former Yugoslav Republic of Macedonia", "The former Yugoslav republic of Macedonia")

LifeExp <- left_join(LifeExp,adlt_mort, by=c("Country","Year"))
LifeExp <- LifeExp %>% 
  dplyr::mutate("Adult.Mortality" = Adult.Mort2)
#drop new variables
drop = c("GDP2","Schooling2","comp2", "Population2","Adult.Mort2")
LifeExp <- LifeExp[,!colnames(LifeExp) %in% drop]
LifeExpKNN = LifeExp


```

### Full Correlation Matrix for Linear Regression (Life.expectancy)
```{r fig.dim = c(14, 12), fig.align='center',out.extra='angle=90'}
#####################################################################################
#      Full Correlation Matrix for Linear Regression (Life.expectancy)              #
#####################################################################################
# Filter for data to be included
num_cols = LifeExp %>% dplyr::select(where(is.numeric)) %>% colnames()
LifeExpcorr = LifeExp[,num_cols]
corrplot(cor(LifeExpcorr), method = 'square', order = 'AOE', addCoef.col = 'black', 
         cl.pos = 'n', col = COL2('BrBG'))

plot_correlate(LifeExp)

```

### Observations
* Under five death and infant death are perfectly correlated. They are describing the same thing. One of the variable is redundant.
* GDP and percentage expenditure are perfectly correlated. They are describing the same thing. One of the variable is redundant.
* Schooling and Income Composition of Resources have a strong positive correlation
* Life Expectancy and Adult Mortality are highly negatively correlated
* Life Expectancy and HIV.AIDS are moderately correlated
* Life Expectancy and BMI are moderately correlated
* Life Expectancy and Schooling are highly correlated
* Life Expectancy and Income Composition of Resources are highly correlated

## Uni-variate analysis
```{r,fig.align='center',out.extra='angle=90'}
#####################################################################################
#                               Uni-variate analysis                                #
#####################################################################################
# Let's plot the summary statistics
# Univariate analysis
num_cols = LifeExp %>% dplyr::select(where(is.numeric)) %>% colnames()
num_cols_exclude = c('Year')
num_cols_plots = noquote(unlist(num_cols[!( num_cols %in% num_cols_exclude)]))
nrows = length(num_cols_plots)
for (i in num_cols_plots)
{
box_p = LifeExp %>%
  ggplot(aes(x="", y = .data[[i]])) +
  geom_boxplot(fill = "sandybrown", color = "black") + 
  coord_flip() + theme_classic() + xlab("") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  ylab(i)
hist_p = LifeExp %>%
  ggplot() +
  geom_histogram(aes(x = .data[[i]], y = (..count..)/sum(..count..)),
                 position = "identity", bins = 30, 
                 fill = "sandybrown", color = "black") +
  ylab("Relative Frequency") +
  theme_classic() + xlab(i) + ggtitle(paste(i, "- Univariate Analysis")) + 
  theme(plot.title = element_text(hjust = 0.5))
egg::ggarrange(hist_p, box_p, heights = 2:1) 
}

```

### Categorical data plots
```{r,fig.align='center',out.extra='angle=90', fig.dim = c(8, 6)}
#####################################################################################
#                               Categorical data plots                              #
#####################################################################################
num_var = LifeExp %>% dplyr::select(where(is.numeric)) %>% colnames()
cat_cols = LifeExp %>% dplyr::select(where(is.factor)) %>% colnames()
num_ex = c('Year')
num_var_plots = noquote(unlist(num_var[!( num_var %in% num_ex)]))
# Plot all categorical variables
for (c in cat_cols)
{
  cat_plot = LifeExp %>% ggplot(aes(x= .data[[c]], group = 1)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..),
                   y= ..prop.. ), stat= "count", vjust = -.5) +
    labs(y = "Percent") +
    scale_y_continuous(labels = scales::percent) + theme(legend.position = "none") +
    ggtitle(paste(c, "Categorical Analysis")) + 
    theme(plot.title = element_text(hjust = 0.5)) + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) # +
    #scale_fill_brewer(palette="Oranges")
    egg::ggarrange(cat_plot, ncol=2) 
}
```

## Bi-variate analysis with Status variable
```{r,fig.align='center',out.extra='angle=90'}
#####################################################################################
#                     Bi-variate analysis with Status variable                      #
#####################################################################################
for (i in num_var_plots)
{
multibox = LifeExp %>%
  ggplot(aes(x=Status, y = .data[[i]])) +
  geom_boxplot(fill = "sandybrown", color = "black") + 
  xlab("Status") +
  ylab(i) + stat_summary(fun=mean, geom="point", shape=20, size=7, color="red", fill="red") +
  ggtitle(paste(i, "vs Status bi-variate analysis")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_brewer(palette = "Oranges")  
egg::ggarrange(multibox, ncol=2)
}
```

## Bi-variate analysis with Region variable
```{r,fig.align='center',out.extra='angle=90'}
#####################################################################################
#                     Bi-variate analysis with Region variable                   #
#####################################################################################
for (i in num_var_plots)
{
multibox = LifeExp %>%
  ggplot(aes(x=Region, y = .data[[i]])) +
  geom_boxplot(fill = "sandybrown", color = "black") + 
  xlab("Region") +
  ylab(i) + stat_summary(fun=mean, geom="point", shape=20, size=7, color="red", fill="red") +
  ggtitle(paste(i, "vs Region bi-variate analysis")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
    scale_fill_brewer(palette = "Oranges")  
egg::ggarrange(multibox, ncol=2)
}
```
## Overall life expectancy over time
```{r,fig.align='center',out.extra='angle=90'}
#####################################################################################
#                          Overall life expectancy over time                        #
#####################################################################################

mean_LifeExp = LifeExp %>% group_by(Year) %>% summarise_at(vars(Life.expectancy), list(meanle = mean))
mean_LifeExp_reg = LifeExp %>% group_by(Year, Region) %>% summarise_at(vars(Life.expectancy), list(meanle = mean))

#Overall life expectancy
ggplot(data=mean_LifeExp, aes(x=Year, y=meanle)) +
  geom_line()+
  geom_point() +
   ggtitle("Mean life expactancy by year") +
  theme(plot.title = element_text(hjust = 0.5))+
   xlab("Year") + ylab("Average Life Expectancy")
  

#Overall life expectancy by Region
ggplot(data=mean_LifeExp_reg, aes(x=Year, y=meanle, group = Region)) +
  geom_line(aes(color=Region))+
  geom_point(aes(color=Region)) +
  ggtitle("Mean life expactancy by year by region") +
theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Year") + ylab("Average Life Expectancy")
```


### Categorical data plots
```{r,fig.align='center',out.extra='angle=90', fig.dim = c(8, 6)}
#####################################################################################
#                               Categorical data plots                              #
#####################################################################################
num_var = LifeExp %>% dplyr::select(where(is.numeric)) %>% colnames()
cat_cols = LifeExp %>% dplyr::select(where(is.factor)) %>% colnames()
num_ex = c('Year')
num_var_plots = noquote(unlist(num_var[!( num_var %in% num_ex)]))
# Plot all categorical variables
for (c in cat_cols)
{
  cat_plot = LifeExp %>% ggplot(aes(x= .data[[c]], group = 1)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..),
                   y= ..prop.. ), stat= "count", vjust = -.5) +
    labs(y = "Percent") +
    scale_y_continuous(labels = scales::percent) + theme(legend.position = "none") +
    ggtitle(paste(c, "Categorical Analysis")) + 
    theme(plot.title = element_text(hjust = 0.5)) + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) # +
    #scale_fill_brewer(palette="Oranges")
    egg::ggarrange(cat_plot, ncol=2) 
}
```

## Bi-variate analysis with Status variable
```{r,fig.align='center',out.extra='angle=90'}
#####################################################################################
#                     Bi-variate analysis with Status variable                      #
#####################################################################################
for (i in num_var_plots)
{
multibox = LifeExp %>%
  ggplot(aes(x=Status, y = .data[[i]])) +
  geom_boxplot(fill = "sandybrown", color = "black") + 
  xlab("Status") +
  ylab(i) + stat_summary(fun=mean, geom="point", shape=20, size=7, color="red", fill="red") +
  ggtitle(paste(i, "vs Status bi-variate analysis")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_brewer(palette = "Oranges")  
egg::ggarrange(multibox, ncol=2)
}
```

## Bi-variate analysis with Region variable
```{r,fig.align='center',out.extra='angle=90'}
#####################################################################################
#                     Bi-variate analysis with Region variable                   #
#####################################################################################
for (i in num_var_plots)
{
multibox = LifeExp %>%
  ggplot(aes(x=Region, y = .data[[i]])) +
  geom_boxplot(fill = "sandybrown", color = "black") + 
  xlab("Region") +
  ylab(i) + stat_summary(fun=mean, geom="point", shape=20, size=7, color="red", fill="red") +
  ggtitle(paste(i, "vs Region bi-variate analysis")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
    scale_fill_brewer(palette = "Oranges")  
egg::ggarrange(multibox, ncol=2)
}
```
## Overall life expectancy over time
```{r,fig.align='center',out.extra='angle=90'}
#####################################################################################
#                          Overall life expectancy over time                        #
#####################################################################################

mean_LifeExp = LifeExp %>% group_by(Year) %>% summarise_at(vars(Life.expectancy), list(meanle = mean))
mean_LifeExp_reg = LifeExp %>% group_by(Year, Region) %>% summarise_at(vars(Life.expectancy), list(meanle = mean))

#Overall life expectancy
ggplot(data=mean_LifeExp, aes(x=Year, y=meanle)) +
  geom_line()+
  geom_point() +
   ggtitle("Mean life expactancy by year") +
  theme(plot.title = element_text(hjust = 0.5))+
   xlab("Year") + ylab("Average Life Expectancy")
  

#Overall life expectancy by Region
ggplot(data=mean_LifeExp_reg, aes(x=Year, y=meanle, group = Region)) +
  geom_line(aes(color=Region))+
  geom_point(aes(color=Region)) +
  ggtitle("Mean life expactancy by year by region") +
theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Year") + ylab("Average Life Expectancy")
```


```{r transform variables in the dataframe}
#####################################################################################
#                                     Data Preparation                              #
#####################################################################################

### Transform variables
LifeExp = mutate(LifeExp, log.HIV.AIDS = ifelse(HIV.AIDS==0, log(HIV.AIDS+1), log(HIV.AIDS)))
LifeExp = mutate(LifeExp, log.GDP = ifelse(GDP==0, log(GDP+1), log(GDP)))
LifeExp = mutate(LifeExp, log.percentage.expenditure = ifelse(percentage.expenditure==0, log(percentage.expenditure+1), log(percentage.expenditure)))


### Transform countries to continents
#library(countrycode)
#LifeExp$Continent = countrycode(sourcevar = LifeExp[, "Country"], origin = "country.name", destination = "continent")
#LifeExp$Continent = as.factor(LifeExp$Continent)
#LifeExp$Status = as.factor(LifeExp$Status)


drop = c('HIV.AIDS', 'percentage.expenditure', 'GDP')#keeping country in
LifeExp = LifeExp[, !(colnames(LifeExp) %in% drop)]


```



```{r train test set}
#####################################################################################
#                    Split the Data to Train and Test sets (85%-15%)               #
#####################################################################################
index<-sample(1:dim(LifeExp)[1],round(dim(LifeExp)[1]*0.85),replace=F)
train = LifeExp[index,]
test = LifeExp[-index,]
```



```{r EDA on Training Data}
#####################################################################################
#                             EDA on Train sets                                    #
#####################################################################################

describe(train)
```

## Observations

** A High degree of skewness can be identified in Infant Deaths
** A High degree of skewness can be identified in Measles
** A High degree of skewness can be identified in under.five.deaths	
** A High degree of skewness can be identified in Population


```{r normality}
normality(train) 
#Runs a Shapario-Wilk Tests, if the p-value is >= .05 then the data is normally distrusted, if <0.05 the data is not normally distrusted.

#Find Features that are not normally distributed 

train %>%
  normality() %>%
  filter(p_value < 0.05) %>%
  arrange(abs(p_value))

# Verify non normality and transformation options of the variability 
plot_normality(train)

```

## Observations

** The p-value of the Population is less than 0.0001 signifying it is non normally and should be transformed if utilized in the following models.	
** The p-value of the Measles is less than 0.0001 signifying it is non normally and should be transformed if utilized in the following models.
** The p-value of the infant.deaths is less than 0.0001 signifying it is non normally and should be transformed if utilized in the following models.
** The p-value of the under.five.deaths is less than 0.0001 signifying it is non normally and should be transformed if utilized in the following models.
** The p-value of the Hepatitis.B is less than 0.0001 signifying it is non normally and should be transformed if utilized in the following models.
** The p-value of the Diphtheria is less than 0.0001 signifying it is non normally and should be transformed if utilized in the following models.
** The p-value of the Polio is less than 0.0001 signifying it is non normally and should be transformed if utilized in the following models.
** The p-value of the log.HIV.AIDS is less than 0.0001 signifying it is non normally and should be transformed if utilized in the following models.
** The p-value of the thinness.5.9.years is less than 0.0001 signifying it is non normally and should be transformed if utilized in the following models.
** The p-value of the thinness..1.19.years is less than 0.0001 signifying it is non normally and should be transformed if utilized in the following models.



#####################################################################################
#                                        Modeling                                   #
#####################################################################################

```{r Model building}
#Prediction function
predict.regsubsets =function (object , newdata ,id ,...){
  form=as.formula (object$call [[2]])
  mat=model.matrix(form ,newdata )
  coefi=coef(object ,id=id)
  xvars=names(coefi)
  mat[,xvars]%*%coefi
}

#####################################################################################
#                                       Lasso                                       #
#####################################################################################
library(glmnet)
x=model.matrix(Life.expectancy~.,train)[,-1]
y=train$Life.expectancy
xtest = model.matrix(Life.expectancy~.,test)[,-1]
ytest = test$Life.expectancy

grid=10^seq(10,-2, length =100)
lasso.mod=glmnet(x,y,alpha=1, lambda =grid)
cv.out=cv.glmnet(x,y,alpha=1)
plot(cv.out)

bestlambda = cv.out$lambda.min  #Optimal penalty parameter.  You can make this call visually.
lasso.pred=predict(lasso.mod ,s=bestlambda ,newx=xtest)

testMSE_LASSO<-mean((ytest-lasso.pred)^2)
testMSE_LASSO
coef(lasso.mod,s=bestlambda)

lasso_residuals = (ytest - lasso.pred)
hist(lasso_residuals, main = "Histogram of Residuals")
plot(lasso_residuals, main = "Residuals plot") 
abline(h=0, col="blue")

# Metrics RMSE; R-squared; MAE
postResample(pred = lasso.pred, obs = ytest)

##### Fit Linear Model based on LASSO regularization without factors to measure VIF####
fit.lasso.lm = lm(Life.expectancy ~ Year + Adult.Mortality + Alcohol + log.percentage.expenditure + Hepatitis.B + Measles + BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + log.HIV.AIDS + thinness.5.9.years + Income.composition.of.resources + Schooling + log.GDP + Population, data = train)
summary(fit.lasso.lm)

### Visualize VIF
fit.lasso.lm_VIF = vif(fit.lasso.lm)
barplot(fit.lasso.lm_VIF, main = 'VIF Values', horiz = TRUE, col="blue", xlim = c(0,12))
abline(v=10, col="red")


######################################################################################################
#Tamas I believe that with the new data set there is no colinearity based on VIF unless we go with 4 #
# I'm commenting out the next section of code for now                                                #
######################################################################################################


# We can see that LASSO did not remove infant.deaths or under.five.deaths as those are perfectly correlated. Let's remove the one with the smallest coefficient (infant.death).

##### Fit Linear Model based on LASSO regularization without factors to measure VIF####
#fit.lasso.lm2 = lm(Life.expectancy ~ Year + Adult.Mortality + Alcohol + Hepatitis.B + Measles + BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + thinness.5.9.years + Income.composition.of.resources + Schooling + log.HIV.AIDS + log.GDP + log.percentage.expenditure, data = train)
#summary(fit.lasso.lm2)

### re-run Visualize VIF
#fit.lasso.lm_VIF2 = vif(fit.lasso.lm2)
#barplot(fit.lasso.lm_VIF2, main = 'VIF Values', horiz = TRUE, col="blue", xlim = c(0,12))
#abline(v=10, col="red")

##### Fit Linear Model based on LASSO regularization and removed multicollinearity####

##### Continent and Region are correlated, dropped Continent for now, because the lasso coef was closer to 0###

fit.lasso.lm3 = lm(Life.expectancy ~ Year + Status + Adult.Mortality + Alcohol + log.percentage.expenditure + Hepatitis.B + Measles + BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + log.HIV.AIDS + thinness.5.9.years + Income.composition.of.resources + Schooling + Region + log.GDP+Population, data = train)

#### Hypothesis testing ####
summary(fit.lasso.lm3)
# At alpha = 0.05 the following variables are not significant therefore don't contribute to the model performance:
# Alcohol, Measles, BMI, Total.expenditure.

# Predicting
train_pred = predict(fit.lasso.lm3, train)
test_pred = predict(fit.lasso.lm3, test)

# Scoring the final model on Training and Test set
residuals = resid(fit.lasso.lm3)
postResample(pred = train_pred, obs = train$Life.expectancy)
postResample(pred = test_pred, obs = test$Life.expectancy)
sm = summary(fit.lasso.lm3)
mse = mean(sm$residuals^2)

### Checking Multiple Liner Regression model assumptions
confint(fit.lasso.lm3)
hist(residuals, main = "Histogram of Residuals")
plot(residuals, main = "Residuals plot") 
abline(h=0, col="blue")
plot(fit.lasso.lm3, which = 2)
plot(fit.lasso.lm3, which = 4)

##### Visualize prediction vs actual
x_lasso = 1:dim(xtest)[1]
plot(x_lasso, ytest, col = "red", type = "l", lwd=2,
     main = "Life Expectancy prediction (LASSO)", ylab="Life expectancy")
lines(x_lasso, test_pred, col = "blue", lwd=2)
legend("topright",  legend = c("original observation", "predicted life expectancy"), 
       fill = c("red", "blue"), col = 2:3,  adj = c(0, 0.6))
grid()

#####################################################################################
#                                  Forward Selection                                #
#####################################################################################
library(leaps)
mlr.fwd=regsubsets(Life.expectancy~.,data=train,method="forward",nvmax=41)
testASE<-c()
for (i in 1:41){
  predictions = predict.regsubsets(object=mlr.fwd,newdata=test,id=i) 
  testASE[i] = mean((test$Life.expectancy-predictions)^2)
}
par(mfrow=c(1,1))
plot(1:41,testASE,type="l",xlab="# of predictors",ylab="test vs train ASE")
index<-which(testASE==min(testASE))
points(index,testASE[index],col="red",pch=10)
rss<-summary(mlr.fwd)$rss
lines(1:41,rss/dim(train)[1],lty=3,col="blue")  

mlr.fwd.final=regsubsets(Life.expectancy~.,data=LifeExp,method="forward",nvmax=41)
coef(mlr.fwd.final,37)
summary(mlr.fwd.final)


# Metrics RMSE; R-squared; MAE
postResample(pred = predictions, obs = ytest)

##### Fit Linear Model based on Forward Selection without factors to measure VIF####
fit.fwd.lm = lm(Life.expectancy ~ Year+Adult.Mortality+infant.deaths+Alcohol+Measles+BMI+under.five.deaths+Polio+Total.expenditure+Diphtheria+Population+thinness.5.9.years+Income.composition.of.resources+Schooling+log.HIV.AIDS+log.GDP+log.percentage.expenditure, data = train)


### Visualize VIF
fit.fwd.lm_VIF = vif(fit.fwd.lm)
barplot(fit.fwd.lm_VIF, main = 'VIF Values', horiz = TRUE, col="blue", xlim = c(0,12))
abline(v=10, col="red")


# We can see that Forward Selection did not remove infant.deaths or under.five.deaths as those are perfectly correlated. Let's remove the one with the smallest coefficient (under.five.deaths).

##### Fit Linear Model based on FWD selection without factors to measure VIF####
fit.fwd.lm2 = lm(Life.expectancy ~ Year+Adult.Mortality+infant.deaths+Alcohol+Measles+BMI+Polio+Total.expenditure+Diphtheria+Population+thinness.5.9.years+Income.composition.of.resources+Schooling+log.HIV.AIDS+log.GDP+log.percentage.expenditure, data = train)
summary(fit.fwd.lm2)

### re-run Visualize VIF
fit.fwd.lm2_VIF = vif(fit.fwd.lm2)
barplot(fit.fwd.lm2_VIF, main = 'Re-test of VIF Values', horiz = TRUE, col="blue", xlim = c(0,12))
abline(v=10, col="red")

##### Fit Linear Model based on Forward Selection regularization and removed multicollinearity and categorical variables####
fit.fwd.lm3 = lm(Life.expectancy ~ Year+Adult.Mortality+Region+Status+infant.deaths+Alcohol+Measles+BMI+Polio+Total.expenditure+Diphtheria+Population+thinness.5.9.years+Income.composition.of.resources+Schooling+log.HIV.AIDS+log.GDP+log.percentage.expenditure, data = train)

#### Hypothesis testing ####
summary(fit.fwd.lm3)
# At alpha = 0.05 the following variables are not significant therefore don't contribute to the model performance:
# thinness.5.9.years, Measles, BMI, Total.expenditure.

# Predicting
train_pred = predict(fit.fwd.lm3, train)
test_pred = predict(fit.fwd.lm3, test)

# Scoring the final model on Training and Test set
residuals = resid(fit.fwd.lm3)
postResample(pred = train_pred, obs = train$Life.expectancy)
postResample(pred = test_pred, obs = test$Life.expectancy)
sm = summary(fit.fwd.lm3)
mse = mean(sm$residuals^2)

### Checking Multiple Liner Regression model assumptions
confint(fit.fwd.lm3)
hist(residuals, main = "Histogram of Residuals")
plot(residuals, main = "Residuals plot") 
abline(h=0, col="blue")
plot(fit.fwd.lm3, which = 2)
plot(fit.fwd.lm3, which = 4)

##### Visualize prediction vs actual
x_fwd = 1:dim(xtest)[1]
plot(x_fwd, ytest, col = "red", type = "l", lwd=2,
     main = "Life Expectancy prediction (FWD selection)", ylab="Life expectancy")
lines(x_fwd, test_pred, col = "blue", lwd=2)
legend("topright",  legend = c("original observation", "predicted life expectancy"), 
       fill = c("red", "blue"), col = 2:3,  adj = c(0, 0.6))
grid()

#####################################################################################
#                               Backward Elimination                                #
#####################################################################################

mlr.bck=regsubsets(Life.expectancy~.,data=train,method="backward",nvmax=41)
testASE<-c()
for (i in 1:41){

  predictions = predict.regsubsets(object=mlr.bck,newdata=test,id=i) 
  testASE[i] = mean((test$Life.expectancy-predictions)^2)
}
par(mfrow=c(1,1))

plot(1:41,testASE,type="l",xlab="# of predictors",ylab="test vs train ASE")
index<-which(testASE==min(testASE))
points(index,testASE[index],col="red",pch=10)
rss<-summary(mlr.bck)$rss
lines(1:41,rss/dim(train)[1],lty=3,col="blue")  

mlr.bck.final=regsubsets(Life.expectancy~.,data=LifeExp,method="backward",nvmax=41)
coef(mlr.bck.final,35)

summary(mlr.bck.final)

# Metrics RMSE; R-squared; MAE
postResample(pred = predictions, obs = ytest)

##different result now, will repeat steps from earlier model

fit.bck.lm <- lm(Life.expectancy ~ Year+Adult.Mortality+infant.deaths+Alcohol+Measles+under.five.deaths+Polio+Total.expenditure+Diphtheria+Population+thinness.5.9.years+Income.composition.of.resources+Schooling+log.HIV.AIDS+log.GDP+log.percentage.expenditure,data = train)

summary(fit.bck.lm)

### Visualize VIF
fit.bck.lm_VIF = vif(fit.bck.lm)
barplot(fit.bck.lm_VIF, main = 'VIF Values', horiz = TRUE, col="blue", xlim = c(0,12))
abline(v=10, col="red")

#two are correlated and giving a high VIF it's under.five and infant.deaths
fit.bck.lm2 <- lm(Life.expectancy ~ Year+Adult.Mortality+Alcohol+Measles+under.five.deaths+Polio+Total.expenditure+Diphtheria+Population+thinness.5.9.years+Income.composition.of.resources+Schooling+log.HIV.AIDS+log.GDP+log.percentage.expenditure,data = train)

fit.bck.lm2_VIF <- vif(fit.bck.lm2)
barplot(fit.bck.lm2_VIF, main = 'VIF Values', horiz = TRUE, col="blue", xlim = c(0,12))
abline(v=10, col="red")

##fit model with categorical variables and removed colinearity
fit.bck.lm3 <- lm(Life.expectancy ~ Year+Adult.Mortality+Alcohol+Measles+under.five.deaths+Polio+Total.expenditure+Diphtheria+Population+thinness.5.9.years+Income.composition.of.resources+Schooling+log.HIV.AIDS+log.GDP+log.percentage.expenditure + Status + Region,data = train)


#### Hypothesis testing ####
summary(fit.bck.lm3)
# At alpha = 0.05 the following variables are not significant therefore don't contribute to the model performance:
# BMI, Total.expenditure, thinness.5.9.years

# Predicting
train_pred = predict(fit.bck.lm3, train)
test_pred = predict(fit.bck.lm3, test)

# Scoring the final model on Training and Test set
residuals = resid(fit.bck.lm3)
postResample(pred = train_pred, obs = train$Life.expectancy)
postResample(pred = test_pred, obs = test$Life.expectancy)
sm = summary(fit.bck.lm3)
mse = mean(sm$residuals^2)

### Checking Multiple Liner Regression model assumptions
confint(fit.bck.lm3)
hist(residuals, main = "Histogram of Residuals")
plot(residuals, main = "Residuals plot") 
abline(h=0, col="blue")
plot(fit.fwd.lm3, which = 2)
plot(fit.fwd.lm3, which = 4)

#####################################################################################
#                                  Ridge regression                                 #
#####################################################################################
x=model.matrix(Life.expectancy~.,train)[,-1]
y=train$Life.expectancy
xtest = model.matrix(Life.expectancy~.,test)[,-1]
ytest = test$Life.expectancy

grid=10^seq(10,-2, length =100)
ridge.mod=glmnet(x,y,alpha=0, lambda =grid, family = 'gaussian') # alpha is 0 for Ridge
cv.out=cv.glmnet(x,y,alpha=0)
plot(cv.out)

bestlambda = cv.out$lambda.min  #Optimal penalty parameter.  You can make this call visually.
ridge.pred=predict(ridge.mod ,s=bestlambda ,newx=xtest)

testMSE_RIDGE<-mean((ytest-ridge.pred)^2)
testMSE_RIDGE
coef(ridge.mod,s=bestlambda)

# Metrics RMSE; R-squared; MAE
postResample(pred = ridge.pred, obs = ytest)

##### Visualize prediction vs actual
x_ridge = 1:dim(xtest)[1]
plot(x_fwd, ytest, col = "red", type = "l", lwd=2,
     main = "Life Expectancy prediction (Ridge Regression)", ylab="Life expectancy")
lines(x_fwd, ridge.pred, col = "blue", lwd=2)
legend("topright",  legend = c("original observation", "predicted life expectancy"), 
       fill = c("red", "blue"), col = 2:3,  adj = c(0, 0.6))
grid()


#####################################################################################
#                            Elastic Net Regression                                 #
#####################################################################################

library(glmnetUtils)
cva.out = cva.glmnet(x,y)
plot(cva.out)

alpha = cva.out$alpha
mse = sapply(cva.out$modlist, function(mod) {min(mod$cvm)})
lambdaMin <- sapply(cva.out$modlist, `[[`, "lambda.min")
min_mse <- which.min(mse)
cva.min = data.frame(alpha = alpha[min_mse], lambdaMin = lambdaMin[min_mse], mse = mse[min_mse])
cva.min

elastic.mod = glmnet(x,y, alpha = cva.min$alpha, lambda = cva.min$lambdaMin)
elastic.pred=predict(elastic.mod ,s=cva.min$lambdaMin ,newx=xtest)
elastic.pred_coef=predict(elastic.mod ,s=cva.min$lambdaMin ,newx=xtest, type = "coef")

testMSE_ELASTIC<-mean((ytest-elastic.pred)^2)
testMSE_ELASTIC

# Metrics RMSE; R-squared; MAE
postResample(pred = elastic.pred, obs = ytest)

##### Visualize prediction vs actual
x_eNET = 1:dim(xtest)[1]
plot(x_eNET, ytest, col = "red", type = "l", lwd=2,
     main = "Life Expectancy prediction (ElasticNet Regression)", ylab="Life expectancy")
lines(x_fwd, elastic.pred, col = "blue", lwd=2)
legend("topright",  legend = c("original observation", "predicted life expectancy"), 
       fill = c("red", "blue"), col = 2:3,  adj = c(0, 0.6))
grid()

#####################################################################################
#                                      Manual MLR - Tamas                           #
#####################################################################################
# 5-fold cross validation
cv <- trainControl(
  method = "cv", 
  number = 5,
  savePredictions = TRUE
)

MLRT = train(
  Life.expectancy ~ Status + Region + Income.composition.of.resources + Schooling + log.HIV.AIDS + log.percentage.expenditure + Year + Adult.Mortality + infant.deaths,
  data = train,
  method = "lm",
   preProcess = c("center", "scale"),
  trControl = cv)

### Visualize VIF
MLR_VIF = vif(MLRT$finalModel)
barplot(MLR_VIF, main = 'VIF Values', horiz = TRUE, col="blue", xlim = c(0,12))
abline(v=10, col="red")

### Hypothesis testing
summary(MLRT$finalModel)

# Predicting
train_pred = predict(MLRT, train)
test_pred = predict(MLRT, test)

# Scoring the final model on Training and Test set
summary(MLRT$finalModel)
residuals = resid(MLRT$finalModel)
postResample(pred = train_pred, obs = train$Life.expectancy)
postResample(pred = test_pred, obs = test$Life.expectancy)

### Checking Multiple Liner Regression model assumptions
fit = lm(Life.expectancy ~ Status + Region + Income.composition.of.resources + Schooling + log.HIV.AIDS + log.percentage.expenditure + Year + Adult.Mortality + infant.deaths, data = train)
confint(fit)
hist(residuals, main = "Histogram of Residuals")
plot(residuals, main = "Residuals plot") 
abline(h=0, col="blue")
plot(fit, which = 2)
plot(fit, which = 4)


##### Visualize prediction vs actual
x_TMLR = 1:dim(xtest)[1]
plot(x_TMLR, ytest, col = "red", type = "l", lwd=2,
     main = "Life Expectancy prediction (Manual MLR)", ylab="Life expectancy")
lines(x_fwd, test_pred, col = "blue", lwd=2)
legend("topright",  legend = c("original observation", "predicted life expectancy"), 
       fill = c("red", "blue"), col = 2:3,  adj = c(0, 0.6))
grid()
```


```{r Objective 2 interaction}
#####################################################################################
#                                      Objective 2                                  #
#                   Model with complexity (adding interaction terms)                #
#####################################################################################
fit_interaction = lm(Life.expectancy ~ Status + Region + Schooling + log.HIV.AIDS + log.percentage.expenditure + Year + Adult.Mortality + infant.deaths + Income.composition.of.resources:Status, train)

#fit_interaction = lm(Life.expectancy ~ Status + Region + Schooling + log.HIV.AIDS + log.percentage.expenditure + Year + Adult.Mortality + infant.deaths + country:infant.mortality.rate, train)

interact_pred_train = predict(fit_interaction, train)
postResample(pred = interact_pred_train, obs = train$Life.expectancy)

interact_pred = predict(fit_interaction, test)
postResample(pred = interact_pred, obs = test$Life.expectancy)

summary(fit_interaction)
confint(fit_interaction)
hist(residuals, main = "Histogram of Residuals")
plot(residuals, main = "Residuals plot") 
abline(h=0, col="blue")
plot(fit_interaction, which = 2)
plot(fit_interaction, which = 4)
```

```{r creating binary categorical variables}
#resplit here for now
index<-sample(1:dim(LifeExp)[1],round(dim(LifeExp)[1]*0.85),replace=F)
train = LifeExp[index,]
test = LifeExp[-index,]
#duplicate dataframe
LifeExp2 <- LifeExp
#create cat variables for HIV and Meas as binary No/Yes
LifeExp2 <- LifeExp2 %>% dplyr::mutate(HIV_cat = if_else(LifeExp2$HIV.AIDS>0.1,1,0),
                                       Meas_cat= if_else(LifeExp2$Measles>0,1,0))

train_custom <- LifeExp2[index,]
test_custom <- LifeExp2[index,]
```
```{r run custom model}
#retrain/fit/test/predict and measure accuracy
fit.custom <- lm(Life.expectancy ~Adult.Mortality + infant.deaths + log.GDP +Measles + HIV_cat + Country + Country:infant.deaths +Country:Adult.Mortality + Adult.Mortality:infant.deaths:Country,data = train_custom)
#model has interactions between HIV and Measles variables and their respective
#binary categories to have a sort of conditional
#interaction between Country:infant.deaths

#check model performance on the test set and test set prediction
test_custom_pred <- predict(fit.custom,test_custom) #rank deficient warning comes from the self-interactions (ex. HIV_cat:HIV)

caret::postResample(pred = test_custom_pred,obs = test_custom$Life.expectancy)
forecast::accuracy(test_custom_pred,test_custom$Life.expectancy)
```

```{r Objective 2 KNN}
#####################################################################################
#                                      Objective 2                                  #
#                                     KNN - regression                              #
#####################################################################################
# Check for zero variance
caret::nearZeroVar(LifeExpKNN %>% dplyr::select(where(is.numeric)), saveMetrics = TRUE) %>% 
  tibble::rownames_to_column() %>% 
  filter(nzv)
cat("No variable with zero variance in the selected list of variables.") 

# Scale the numerical variables as KNN is sensitive to that
#LifeExpKNN_scale = LifeExpKNN %>% mutate_if(is.numeric, scale)

### KNN Data Preparation
index<-sample(1:dim(LifeExpKNN)[1],round(dim(LifeExpKNN)[1]*0.85),replace=F)
KNNtrain = LifeExpKNN[index,]
KNNtest = LifeExpKNN[-index,]
x=KNNtrain[,c(-4, -23)]
y=KNNtrain$Life.expectancy
xtest = KNNtest[,c(-4,-23)]
ytest = KNNtest$Life.expectancy
KNN_drop = c(-23,-12, -1)

# Search for optimal k
k_grid = expand.grid(k = seq(3, 25, by = 2))

# Model Training
KNNRegressor = train(
  Life.expectancy~.,
  data = KNNtrain[,KNN_drop],
  method = "knn",
  preProcess = c("center", "scale"), 
  tuneGrid = k_grid,
  metric     = "RMSE",
  trControl = cv
  )

prediction_test = predict(KNNRegressor, KNNtrain[,KNN_drop])
postResample(pred = prediction_test, obs = KNNtrain[,KNN_drop]$Life.expectancy)

prediction_test = predict(KNNRegressor, KNNtest[,KNN_drop])
postResample(pred = prediction_test, obs = KNNtest[,KNN_drop]$Life.expectancy)

plot(KNNRegressor)
varImp(KNNRegressor)

x = 1:length(KNNtest$Life.expectancy)
plot(x, KNNtest$Life.expectancy, col = "red", type = "l", lwd=2,
     main = "Life Expectancy prediction")
lines(x, prediction_test, col = "blue", lwd=2)
legend("topright",  legend = c("original observation", "predicted life expectancy"), 
       fill = c("red", "blue"), col = 2:3,  adj = c(0, 0.6))
grid()
```

### Visualize 'k' and the most important features
```{r, fig.align='center',out.extra='angle=90'}
# Visualize 'k' and the most important features
ggplot(KNNRegressor) + ggtitle("Optimal k value for the highest accuracy") +
  theme(plot.title = element_text(hjust = 0.5))
KNNvarImp = varImp(KNNRegressor)
plot(KNNvarImp, top = 5, main='Top 5 Variable predicting life expectancy (KNN)')
```